---
title: "Clustering and Dimensionality Reduction"
output: pdf_document
date: "2023-08-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, include=FALSE}
library(tidyverse)
library(ggplot2)
library(caret)
library(flexclust)
library(foreach)
```

```{r, echo=FALSE}
wine = read.csv("../STA380/data/wine.csv") 
wine_chems=wine[,c(1:11)]
wine_chems = scale(wine_chems, center=TRUE, scale=TRUE)	
```

### Let's start with a look at the correlation heatmap

```{r, echo=FALSE}
# a quick heatmap visualizations
ggcorrplot::ggcorrplot(cor(wine_chems))
```

### Let's order them according to their correlations

```{r, echo=FALSE}
# looks a mess -- reorder the variables by hierarchical clustering
ggcorrplot::ggcorrplot(cor(wine_chems), hc.order = TRUE)
```

### We can see a couple of obvious strong correlations
Density and alcohol seem to be negatively correlated, which is expected since alcohol is less dense than water, and increasing alcohol will lead to decrease in quality.

Density and residual sugar seem to be positively correlated, again expected as addign sugar will make a liquid more dense.

```{r, echo=FALSE}
# Now look at PCA of the (average) survey responses.  
# This is a common way to treat survey data
PCAwine = prcomp(wine_chems, scale=FALSE,rank. = 2)
```


```{r, echo=FALSE, include=FALSE}
## variance plot
plot(PCAwine)
summary(PCAwine)
```


### Lets do PCA and reduce the predictors to just 2 Principal Componets

```{r, echo=FALSE}
# first few pcs
# try interpreting the loadings
# the question to ask is: "which variables does this load heavily on (positive and negatively)?"
round(PCAwine$rotation[,1:2],2) 
```


```{r, echo=FALSE}
# create a tidy summary of the loadings
loadings_summary = PCAwine$rotation %>%
  as.data.frame() %>%
  rownames_to_column('Question')
```

PC1 seems to pick out characteristics of sulfury vs vinegary wines.

PC2 just seems to load positively on alcohol which is negatively correlated with density.

### Let's plot the wines based on these principal components and color them with their actul colors

```{r, echo=FALSE}
wine = merge(wine, PCAwine$x[,1:2], by="row.names")
wine = rename(wine, Show = Row.names)

wine$quality_c = as.factor(ifelse(wine$quality>6,1,0))
#Let's plot the wines with their colors
ggplot(wine, aes(x = PC1, y = PC2, color = color)) +
  geom_point(size = 3) +
  labs(title = "Wines with their Actual Color Categories", x = "PC1", y = "PC2")

```

They seem to be well separated

### What about quality though, Let's simplify the plot by converting quality into a binary variable where 1 means high quality ie quality >=7
```{r, echo=FALSE}


ggplot(wine, aes(x = PC1, y = PC2, color = quality_c)) +
  geom_point(size = 3) +
  labs(title = "Wines with their Actual Quality Ratings(1 means 7 or higher)", x = "PC1", y = "PC2")

```

Looks like PCA didn't pick on quality rather color of wines

### If we cluster the wines will they cluster into their own colors?

```{r, echo=FALSE}
wine_c_data=wine[,c('PC1','PC2')]

# Run k-means with 2 clusters
clust1 = kmeans(wine_c_data, 2, nstart=25)

#Since we have more white wines in the data set, let's name bigger cluster as white
wine_c_data$cc = ifelse(clust1$cluster==which.max(clust1$size),"white","red")

#summary(clust1$cluster)
# Create the scatter plot
ggplot(wine_c_data, aes(x = PC1, y = PC2, color = cc)) +
  geom_point(size = 3)  +
  labs(title = "Wines with their predicted Color Categories", x = "PC1", y = "PC2")
```

### Seems like they do. but how well, let's take a look at the confusion matrix

```{r, echo=FALSE}
confusionMatrix(as.factor(wine_c_data$cc),as.factor(wine$color))
```

### Wow 98% of the wines did find their own kind

## Let's try the same using tSNE now

```{r, include= FALSE}
library(tidyverse)
library(ggplot2)
library(Rtsne)
library(flexclust)
library(foreach)
library(caret)
```


```{r, include=FALSE}
set.seed(1)
wine = read.csv("../STA380/data/wine.csv") 
wine_d=wine[!duplicated(wine[, 1:11]), ]
wine_chems=wine_d[,c(1:11)]

# Perform t-SNE
tsne_result <- Rtsne(wine_chems, perplexity = 50, dims = 2, verbose = TRUE)
```

### Let's see what our tSNE plot looks like with 2 components
```{r,echo=FALSE}
# Plot the t-SNE result
plot(tsne_result$Y, col = "blue", pch = 20, main = "t-SNE Result")

# create a tidy summary of the loadings
tSNEwine = tsne_result$Y %>%
  as.data.frame()

tSNEwine$quality_c = as.factor(ifelse(wine_d$quality>6,1,0))
```

### Let's plot the wines with their actual colors

```{r,echo=FALSE}
#Let's plot the wines with their colors
ggplot(tSNEwine, aes(x = V1, y = V2, color = wine_d$color)) +
  geom_point(size = 3) +
  labs(title = "Wines with their Actual Color Categories", x = "V1", y = "V2")
```

There seems to be some differentiation but colors are bleeding into each other

### What about quality though

```{r,echo=FALSE}
ggplot(tSNEwine, aes(x = V1, y = V2, color = quality_c)) +
  geom_point(size = 3) +
  labs(title = "Wines with their Actual Quality Ratings(1 means 7 or higher)", x = "V1", y = "V2")
```

Muddy again.

### Let's try K-means clustering to see if the machine can pick out the 2 colors

```{r,echo=FALSE}
wine_c_data=tSNEwine[,c('V1','V2')]

# Run k-means with 2 clusters
clust1 = kmeans(wine_c_data, 2,nstart=25)

which.max(clust1$size)

#Since we have more white wines in the data set, let's name bigger cluster as white
wine_c_data$cc = ifelse(clust1$cluster==which.max(clust1$size),"white","red")

summary(clust1$cluster)
# Create the scatter plot
ggplot(wine_c_data, aes(x = V1, y = V2, color = cc)) +
  geom_point(size = 3)  +
  labs(title = "Wines with their predicted Color Categories", x = "V1", y = "V2")
```

Seems like K means doesn't know it's wines this time.

### How bad is the accuracy?

```{r,echo=FALSE}
confusionMatrix(as.factor(wine_c_data$cc),as.factor(wine_d$color))

```

A bare 42%, I won't be trusting tSNE + K Means for my wines, for sure.


### In conclusion PCA did pretty well on picking out the colors of the wines, but not on the quality.

### tSNE on the other hand did bad on both of them, and picked up on some yet known differentiator between the wines.
